Variational theory of non-relativistic quantum electrodynamics

### 2019-01-29 paper4pg.tex
modifications to text

Motivated by all of these potential advantages, we now develop an ansatz in which the ground state can be considered as a factorizable state of effective matter and effective photon quasiparticles, both in their respective vacuum states. This ansatz $-$ reminiscent to, but qualitatively distinct from, the Hartree-Fock ansatz~\cite{szabo1989} of electronic structure theory $-$ leads to coupled eigen-equations describing ground and excited states of the light-matter system. We apply our ansatz to describe ground and excited states in a multi-level emitter coupled to many photonic modes. We find that for light-matter couplings that respect sum rules, our method yields ground and excited state energies to a remarkable accuracy of up to 99\%, even in deeply non-perturbative coupling regimes.  In regimes where our results are accurate, we have found the effective quasiparticle description of the ground state of QED.  Our findings also furnish a non-perturbative theory of the position-dependent energy (Lamb) shifts of ground and excited states that give rise to Casimir-Polder forces. \textcolor{blue}{The variational method put forth developed in this manuscript is particularly suited for dealing with QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds, and subsequently methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate.} Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.

@@

Importantly, the reason perturbation theory fails for first excited state, much more so than for the ground state, is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly $2$ eV, which is far larger than the spacing of the bare emitter levels. On the other hand, the variational estimate from our formalism finds no such large negative energy shift, and leads to an energy gap between the first two levels which is similar to the bare gap, and in agreement with numerical diagonalization. The reason for this behavior is that the effect of the plasma term in Eq. (\ref{eq:mhf-photon}) is to blue-shift all of the photon frequencies. In particular, for the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition. The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}). This is a so-called light-matter decoupling effect, which was proposed in Ref. ~\cite{liberato2014}. \textcolor{blue}{In Ref. ~\cite{liberato2014}, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, deLiberato obtained "effective field mode profiles" with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}

\textcolor{blue}{This mechanism is also reflected in Fig. ~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle Ap \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in Supplemental Materials). Such observables may play a role in correlated spectroscopies, as proposed in Ref. ~\cite{flick2017}. We show predictions of the value of this observable based on the variational method, numerical diagonalization, and perturbation theory in the bare emitter and photon states. As can be seen, perturbation theory overestimates the magnitude of the energy shift, while both numerical and variational methods capture a saturation and then decrease of this expectation value, showing an apparent de-correlation between light and matter in the deep-strong coupling regime. The results of Fig. ~\ref{fig:results} and \ref{fig:correlated}  demonstrate not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime. The results of Fig. ~\ref{fig:results} and \ref{fig:correlated} also show that despite correlations being treated perturbatively, it remains possible for correlated observables (and energies) to be predicted with high accuracy.}

>>>>

Motivated by all of these potential advantages, we now develop an ansatz in which the ground state can be considered as a factorizable state of effective matter and effective photon quasiparticles, both in their respective vacuum states. This ansatz $-$ reminiscent to, but qualitatively distinct from, the Hartree-Fock ansatz~\cite{szabo1989} of electronic structure theory $-$ leads to coupled eigen-equations describing ground and excited states of the light-matter system. We apply our ansatz to describe ground and excited states in a multi-level emitter coupled to many photonic modes. We find that for light-matter couplings that respect sum rules, our method yields ground and excited state energies to a remarkable accuracy of up to 99\%, even in deeply non-perturbative coupling regimes.  In regimes where our results are accurate, we have found the effective quasiparticle description of the ground state of QED.  Our findings also furnish a non-perturbative theory of the position-dependent energy (Lamb) shifts of ground and excited states that give rise to Casimir-Polder forces. \textcolor{blue}{The variational method developed in this manuscript is particularly suited for dealing with QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds. Thus, methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate in this regime.} Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.

@@

Importantly, the reason perturbation theory fails for first excited state, much more so than for the ground state, is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly $2$ eV, which is far larger than the spacing of the bare emitter levels. On the other hand, the variational estimate from our formalism finds no such large negative energy shift, and leads to an energy gap between the first two levels which is similar to the bare gap, and in agreement with numerical diagonalization. The reason for this behavior is that the effect of the plasma term in Eq. (\ref{eq:mhf-photon}) is to blue-shift all of the photon frequencies. In particular, for the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition. The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}). This is a so-called light-matter decoupling effect, which was proposed in Ref. ~\cite{liberato2014}. \textcolor{blue}{In Ref. ~\cite{liberato2014}, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, the author obtained "effective field mode profiles" with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}

\textcolor{blue}{This mechanism is also reflected in Fig. ~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle A\cdot p \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in Supplemental Materials). Such observables may play a role in correlated spectroscopies, as proposed in Ref. ~\cite{flick2017}. We show predictions of the value of this observable based on the variational method, numerical diagonalization, and perturbation theory in the bare emitter and photon states. As can be seen, perturbation theory overestimates the magnitude of the energy shift, while both numerical and variational methods capture a saturation and then decrease of this expectation value, showing an apparent de-correlation between light and matter in the deep-strong coupling regime. The results of Fig. ~\ref{fig:results} and \ref{fig:correlated}  demonstrate not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime. The results of Fig. ~\ref{fig:results} and \ref{fig:correlated} also show that despite correlations being treated perturbatively, it remains possible for correlated observables (and energies) to be predicted with high accuracy.}
